Experience in Modeling a Microcontroller Instruction Set Using B

### 2012-05-20 trunk/papers/ArticleZ80/source/paper.tex
Versao praticamente final.

We model bits as a set of integers: $\textit{BIT} =
\textit{0..1}$. The negation is an unary function on bits and it is
defined below. The module also provides lemmas on negation that may be useful for the
users of the library to develop proofs.
$$
\begin{array}{l}
\textit{bit\_not}  \in  \textit{BIT}  \fun  \textit{BIT}  \land 
\textit{bit\_not} \rm = \rm \{ \rm 0  \mapsto  \rm 1 , \rm 1  \mapsto  \rm 0 \rm \} \land \\
\forall (\textit{bb}).(\textit{bb} \in \textit{BIT} \implies \textit{bit\_not}(\textit{bit\_not}(\textit{bb})) = \textit{bb})
\end{array}
$$
Conjunction is a binary function on bits and it is defined as:
$$
\begin{array}{l}
\textit{bit\_and} \in \textit{BIT} \times \textit{BIT} \fun \textit{BIT} \land \\
\forall (\textit{b1}, \textit{b2}).(\textit{b1}  \in \textit{BIT}  \land \textit{b2} \in \textit{BIT} \implies \\
\quad ((\textit{bit\_and}(\textit{b1}, \textit{b2}) = 1) \iff (\textit{b1} = 1)  \land  (\textit{b2} = 1)))
\end{array}
$$
%The module provides the following lemmas for commutativity and associativity:
%$$
%\begin{array}{l}
%%  \textit{bit\_and}(0,0) = 0;  \textit{bit\_and}(0,1) = 0; \\
%%  \textit{bit\_and}(1,0) = 0;  \textit{bit\_and}(1,1) = 1; \\
%\forall (\textit{b1},\textit{b2}).(\textit{b1} \in \textit{BIT} \land \textit{b2} \in \textit{BIT} \implies \\
%\quad (\textit{bit\_and}(\textit{b1}, \textit{b2}) = \textit{bit\_and}(\textit{b2},\textit{b1})))\land \\
%\forall (\textit{b1},\textit{b2},\textit{b3}).(\textit{b1} \in \textit{BIT} \land  \textit{b2} \in \textit{BIT} \land \textit{b3} \in \textit{BIT} \implies \\
%\quad (\textit{bit\_and}(\textit{b1}, \textit{bit\_and}(\textit{b2},\textit{b3})) = \textit{bit\_and}(\textit{bit\_and}(\textit{b1},\textit{b2}),\textit{b3})))\\
%% \forall (\textit{b1}).(\textit{b1} \in \textit{BIT} \implies (\textit{bit\_and}(\textit{b1}, 1) = \textit{b1})); \\
%% \forall (\textit{b1}).(\textit{b1} \in \textit{BIT} \implies (\textit{bit\_and}(\textit{b1}, 0) = 0));
%\end{array}
%$$
The module provides lemmas for commutativity and associativity, definitions of $\textit{bit\_or}$ (disjunction)
and $\textit{bit\_xor}$ (exclusive disjunction), as well as lemmas on
those operators. These are standard and their expression in B is
similar to $\textit{bit\_and}$.

@@

We also define two functions $\textit{bv\_set}$ and
$\textit{bv\_clear}$ that, given a bit vector, and a position in a
bit vector, returns the bit vector resulting from setting the
corresponding position to 1 and 0, and a function $\textit{bv\_get}$
that, given a bit vector and a valid position, returns the value of
the bit at that position. Only the first definition is shown here:
$$
\begin{array}{l}
\textit{bv\_set} \in \textit{BIT\_VECTOR} \times \nat \fun \textit{BIT\_VECTOR} \land \\ \textit{bv\_set} =
\lambda v, n . (v \in \textit{BIT\_VECTOR} \land n \in \nat \land n <\textit{size}(v)
\mid v \lover \{ n+1 \mapsto 1 \})
\end{array}
$$

Function $bv\_catenate$ takes as parameters two bit vectors $v$ and
$w$, and returns the concatenation of $v$ and $w$, such
that $v$ constitutes the most significant part of the result.

%
% $
% \begin{array}{l}
% \textit{bv\_catenate} \in \textit{BIT\_VECTOR} \times \textit{BIT\_VECTOR} \fun \textit{BIT\_VECTOR} \land \\
% \textit{bv\_catenate} = \lambda v, w \bullet (v \in \textit{ BIT\_VECTOR} \land w \in \textit{ BIT\_VECTOR}  \mid v 
% %\conc
%   w)
% \end{array}
% $
\begin{flushleft}
\hspace*{0.0in}\it bv\_catenate  $\in$  \it BIT\_VECTOR  $\times$  \it BIT\_VECTOR  $\fun$ \it
BIT\_VECTOR $\land$\\
\it bv\_catenate \rm =  $\lambda$  v\rm ,\it w \rm . \rm (\it v 
$\in$  \it BIT\_VECTOR $\land$ \it w $\in$  \it BIT\_VECTOR  $\mid$  \it v\^ \it w\rm ) %$\cat$
\\

\end{flushleft}

@@

Additionally, the module provides definitions for the classical
logical combinations of bit vectors: $\textit{bit\_not}$,
$\textit{bit\_and}$, $\textit{bit\_or}$ and $\textit{bit\_xor}$. Only
the first two are presented here. The domain of the binary operators
is restricted to pairs of bit vectors of the same length:

\begin{flushleft}
$$
\begin{array}{l}
\textit{bv\_not} \in \textit{BIT\_VECTOR} \fun \textit{BIT\_VECTOR} \land \\
\textit{bv\_not} = \lambda v . (v \in \textit{BIT\_VECTOR} \mid \quad \lambda i . (1 .. \textit{size}(v)) \mid \textit{bit\_not}(v(i))) \land \\
\textit{bv\_and} \in \textit{BIT\_VECTOR} \times \textit{BIT\_VECTOR} \fun \textit{BIT\_VECTOR} \land \\
\textit{bv\_and} = \lambda v_1, v_2 . (v_1 \in \textit{BIT\_VECTOR} \land v_2 \in \textit{BIT\_VECTOR} \land \\
\textit{size}(v_1) = \textit{size}(v_2) \mid \lambda i . (1 .. \textit{size}(v_1)) \mid
\textit{bit\_and}(v_1(i), v_2(i)))
\end{array}
$$
\end{flushleft}

@@

\subsection{Modeling the actions and instructions}
%TODO: Fiz uma breve alteração
Each instruction is represented by a B operation in Z80 module. The main
module (\textit{Z80}) has two categories of operations: a category represents the 
instructions of microcontrollers and a category represents the external actions, e.g.
interruptions actions. All 158 instructions and interruptions were modeled, but only
two important examples are presented below.

@@

\hspace*{0.20in}\bf ext\_update\_io\_ports\rm (\it address\rm ,\it value\rm )\rm =

@@

\hspace*{0.40in}\it io\_ports \rm ( \it uchar\_byte \rm ( \it address \rm ) \rm ) \rm := \it schar\_byte \rm ( \it
value \rm )

\hspace*{0.20in}\bf END\rm

>>>>

We model bits as a set of integers: $\textit{BIT} =
\textit{0..1}$. The negation is an unary function on bits and it is
defined below. The module also provides lemmas on negation that may be useful for the
users of the library to develop proofs.

$$
\begin{array}{l}
\textit{bit\_not}  \in  \textit{BIT}  \fun  \textit{BIT}  \land 
\textit{bit\_not} \rm = \rm \{ \rm 0  \mapsto  \rm 1 , \rm 1  \mapsto  \rm 0 \rm \} \land \\
\forall (\textit{bb}).(\textit{bb} \in \textit{BIT} \implies \textit{bit\_not}(\textit{bit\_not}(\textit{bb})) = \textit{bb})
\end{array}
$$
Conjunction is a binary function on bits and it is defined as:
$$
\begin{array}{l}
\textit{bit\_and} \in \textit{BIT} \times \textit{BIT} \fun \textit{BIT} \land \\
\forall (\textit{b1}, \textit{b2}).(\textit{b1}  \in \textit{BIT}  \land \textit{b2} \in \textit{BIT} \implies \\
\quad ((\textit{bit\_and}(\textit{b1}, \textit{b2}) = 1) \iff (\textit{b1} = 1)  \land  (\textit{b2} = 1)))
\end{array}
$$
%The module provides the following lemmas for commutativity and associativity:
%$$
%\begin{array}{l}
%%  \textit{bit\_and}(0,0) = 0;  \textit{bit\_and}(0,1) = 0; \\
%%  \textit{bit\_and}(1,0) = 0;  \textit{bit\_and}(1,1) = 1; \\
%\forall (\textit{b1},\textit{b2}).(\textit{b1} \in \textit{BIT} \land \textit{b2} \in \textit{BIT} \implies \\
%\quad (\textit{bit\_and}(\textit{b1}, \textit{b2}) = \textit{bit\_and}(\textit{b2},\textit{b1})))\land \\
%\forall (\textit{b1},\textit{b2},\textit{b3}).(\textit{b1} \in \textit{BIT} \land  \textit{b2} \in \textit{BIT} \land \textit{b3} \in \textit{BIT} \implies \\
%\quad (\textit{bit\_and}(\textit{b1}, \textit{bit\_and}(\textit{b2},\textit{b3})) = \textit{bit\_and}(\textit{bit\_and}(\textit{b1},\textit{b2}),\textit{b3})))\\
%% \forall (\textit{b1}).(\textit{b1} \in \textit{BIT} \implies (\textit{bit\_and}(\textit{b1}, 1) = \textit{b1})); \\
%% \forall (\textit{b1}).(\textit{b1} \in \textit{BIT} \implies (\textit{bit\_and}(\textit{b1}, 0) = 0));
%\end{array}
%$$
The module provides lemmas for commutativity and associativity, definitions of $\textit{bit\_or}$ (disjunction)
and $\textit{bit\_xor}$ (exclusive disjunction), as well as lemmas on
those operators. These are standard and their expression in B is
similar to $\textit{bit\_and}$.

@@

We also define two functions $\textit{bv\_set}$ and
$\textit{bv\_clear}$ that, given a bit vector, and a position in a
bit vector, returns the bit vector resulting from setting the
corresponding position to 1 and 0, and a function $\textit{bv\_get}$
that, given a bit vector and a valid position, returns the value of
the bit at that position. Only the first definition is shown here:\newline
%
%\vspace*{4mm}
%\hspace*{-0.2in}\textit{bv\_set} $\in$ \textit{BIT\_VECTOR} $\times \nat \fun$ \textit{BIT\_VECTOR} $\land$\\
%\it bv\_set \rm =  $\lambda$ \rm \it v\rm ,\it n\rm \rm .\rm (\it v $\in$  \it BIT\_VECTOR  $\land$  \it n  $\in$  \rm 0$\upto$\rm (\it size\rm (\it v\rm )\rm -\rm 1\rm )  $\mid$  \it v  $\lover$  \rm \{ \rm (\it n\rm +\rm 1\rm )  $\mapsto$  \rm 1 \rm \}\rm ) 
%\vspace*{4mm}
%
%
% \nat \land n <\textit{size}(v)
$$
\hspace*{0.1in}\begin{array}{l}
\textit{bv\_set} \in \textit{BIT\_VECTOR} \times \nat \fun \textit{BIT\_VECTOR} \land \\
 \textit{bv\_set} = \lambda v, n . (v \in \textit{BIT\_VECTOR} \land n \in   0..\textit{size}(v)-1
  \mid v \lover \{n+1 \mapsto 1\})
\end{array}
$$

Function $bv\_catenate$ takes as parameters two bit vectors $v$ and
$w$, and returns the concatenation of $v$ and $w$, such
that $v$ constitutes the most significant part of the result. \newline

%$$
%\hspace*{-0.2in}\begin{array}{l}
% \textit{bv\_catenate} \in \textit{BIT\_VECTOR} \times \textit{BIT\_VECTOR} \fun \textit{BIT\_VECTOR} \land \\
% \textit{bv\_catenate} = \lambda v, w \bullet (v \in \textit{ BIT\_VECTOR} \land w \in \textit{ BIT\_VECTOR}  \mid v 
% %\conc
%   w)
%\end{array}
%$$

\hspace*{-0.0in}\it bv\_catenate  $\in$  \it BIT\_VECTOR  $\times$  \it BIT\_VECTOR  $\fun$ \it
BIT\_VECTOR $\land$\\
\hspace*{0.2in}\it bv\_catenate \rm =  $\lambda$  v\rm ,\it w \rm . \rm (\it v 
$\in$  \it BIT\_VECTOR $\land$ \it w $\in$  \it BIT\_VECTOR  $\mid$  \it v\^ \it w\rm ) %$\cat$
\newline

@@

Additionally, the module provides definitions for the classical
logical combinations of bit vectors: $\textit{bit\_not}$,
$\textit{bit\_and}$, $\textit{bit\_or}$ and $\textit{bit\_xor}$. Only
the first two are presented here. The domain of the binary operators
is restricted to pairs of bit vectors of the same length:\newline
$$
\hspace*{-0.2in}\begin{array}{l}
\textit{bv\_not} \in \textit{BIT\_VECTOR} \fun \textit{BIT\_VECTOR} \land \\
\textit{bv\_not} = \lambda v . (v \in \textit{BIT\_VECTOR} \mid \quad \lambda i . (1 .. \textit{size}(v)) \mid \textit{bit\_not}(v(i))) \land \\
\textit{bv\_and} \in \textit{BIT\_VECTOR} \times \textit{BIT\_VECTOR} \fun \textit{BIT\_VECTOR} \land \\
\textit{bv\_and} = \lambda v_1, v_2 . (v_1 \in \textit{BIT\_VECTOR} \land v_2 \in \textit{BIT\_VECTOR} \land \\
\textit{size}(v_1) = \textit{size}(v_2) \mid \lambda i . (1 .. \textit{size}(v_1)) \mid
\textit{bit\_and}(v_1(i), v_2(i)))
\end{array}
$$

@@

\subsection{Modeling the actions and instructions}
%TODO: Fiz uma breve alteração
Each instruction is represented by a B operation in Z80 module. The main
module (\textit{Z80}) has two categories of operations: a category represents the 
instructions of microcontrollers and a category represents the external actions, e.g.
interruptions actions. All 158 instructions and interruptions were modeled, but only
two important examples are presented here.

@@

\hspace*{0.00in}\bf ext\_update\_io\_ports\rm (\it address\rm ,\it value\rm )\rm =

@@

\hspace*{0.20in}\it io\_ports \rm ( \it uchar\_byte \rm ( \it address \rm ) \rm ) \rm := \it schar\_byte \rm ( \it
value \rm )
